Reassessing the Role of Theory and Machine Learning in Forecasting Civil Conflict

### 2020-11-10 paper/paper.Rmd
drop small edits into .Rmd

Although their analysis in the end does not support this point, we do not disagree that theory can important in predictive modeling. Of the various ways one could contrast ad hoc machine learning with "theoretical" modeling, B&S focus essentially on theory as a mechanism for variable selection. In this sense, theory is in fact already used, informally, in nominally ad hoc predictive modeling related to conflict forecasting. The problem is that the extent of extant theories for a phenomenon like civil war is so broad and non-specific in terms.  They do not provide much guidance on an exact specification of what variables and which functional form is required. As a result it is both not very clear what the "correct" model should be, nor easy to demonstrate that any particular "theoretical" model's ability to predict generalizes beyond the data that was used to create it in the first place. Rather than trying to contrast theory and machine learning as competing mechanisms for creating predictive models, it would be more productive to focus on how we can use the flexibility of machine learning models and the information predictive modeling may give us about our ability to "explain" a phenomenon to strengthen and whittle down the broad space of extant theories.

@@

Instantiating the theoretical model is more about choosing the correct variables. This is the contrast between theoretical prediction models and atheoretical machine learning that B&S focus on. But this dichotomy is misguided. The data that are used in a predictive model have to come from somewhere. Even if the inclusion of some factors is not explicitly justified, it is hardly the case that one assembles data without the expectation that it will be useful. It is not a coincidence that nobody uses 1,180 variable strawman models whose variable vectors predominantly consist of 0's, even if such a model coincidentally ends up performing well. Rather, we posit that most of the cases B&S might characterize as atheoretical enter the "theoretical", with some awareness of existing arguments and work that is used to select data sources for inclusion in a dataset.\footnote{While we do not undertake this, there are a variety of procedures that one might use for variable choices, ranging from staring at lists of variables to machine learning approaches.}

@@

Conversely, not all factors that are "theoretically" important or suggested by the literature are also important for prediction, and some factors that turn out to be good predictors do not originate from prior theory. For example, both @ward:etal:2010 and @hill:jones:2014 examine a variety of variables that are considered "theoretically" important for explaining civil war onset and state repression, respectively, and find that there is substantial variation in how useful they are for for actually predicting those outcomes. And, it is entirely possible that a focus on building accurate predictive models could uncover factors overlooked in the literature, e.g. the strong association between infant mortality and a broad variety of conflict outcomes like political instability [@goldstone:etal:2010], irregular leadership changes [@beger:etal:2017], and coups [@beger:ward:2020].

@@

Blair & Sambanis have focused attention on comparing forecasting models and forecasts in the civil war domain. As social science becomes more adept at predictive analysis, this will doubtless be of increasing importance. However, these comparisons must be made carefully to ensure those correct inferences are drawn. We continue to think that weak theory is overrated [@ward:2016], and that machine learning and big data will allow us to learn new things. It will be interesting to see how the evidence for these claims is adjudicated with additional usage and careful evaluation. Social science has leaned on theory for (at least) seventy years, but it is time to find useful ways to incorporate recently developed methods in the arena of machine learning and modern statistics.

>>>>

Although their analysis in the end does not support this point, we do not disagree that theory can be important in predictive modeling. Of the various ways one could contrast ad hoc machine learning with "theoretical" modeling, B&S focus essentially on theory as a mechanism for variable selection. In this sense, theory is in fact already used, informally, in nominally ad hoc predictive modeling related to conflict forecasting. The problem is that the extent of extant theories for a phenomenon like civil war is so broad and non-specific in terms.  They do not provide much guidance on an exact specification of what variables and which functional form is required. As a result it is both not very clear what the "correct" model should be, nor easy to demonstrate that any particular "theoretical" model's ability to predict generalizes beyond the data that was used to create it in the first place. Rather than trying to contrast theory and machine learning as competing mechanisms for creating predictive models, it would be more productive to focus on how we can use the flexibility of machine learning models and the information predictive modeling may give us about our ability to "explain" a phenomenon to strengthen and whittle down the broad space of extant theories.

@@

Instantiating the theoretical model is more about choosing the correct variables. This is the contrast between theoretical prediction models and atheoretical machine learning that B&S focus on. But this dichotomy is misguided. The data that are used in a predictive model have to come from somewhere. Even if the inclusion of some factors is not explicitly justified, it is hardly the case that one assembles data without the expectation that it will be useful. It is not a coincidence that nobody uses 1,180 variable strawman models whose variable vectors predominantly consist of 0's, even if such a model coincidentally ends up performing well. Rather, we posit that most of the cases B&S might characterize as atheoretical enter the "theoretical" with some awareness of existing arguments and work that is used to select data sources for inclusion in a dataset.\footnote{While we do not undertake this, there are a variety of procedures that one might use for variable choices, ranging from staring at lists of variables to machine learning approaches.}

@@

Conversely, not all factors that are "theoretically" important or suggested by the literature are also important for prediction, and some factors that turn out to be good predictors do not originate from prior theory. For example, both @ward:etal:2010 and @hill:jones:2014 examine a variety of variables that are considered "theoretically" important for explaining civil war onset and state repression, respectively, and find that there is substantial variation in how useful they are for for actually predicting those outcomes. It is entirely possible that a focus on building accurate predictive models could uncover factors overlooked in the literature, e.g. the strong association between infant mortality and a broad variety of conflict outcomes like political instability [@goldstone:etal:2010], irregular leadership changes [@beger:etal:2017], and coups [@beger:ward:2020].

@@

Blair & Sambanis have focused attention on comparing forecasting models and forecasts in the civil war domain. As social science becomes more adept at predictive analysis, this will doubtless be of increasing importance. However, these comparisons must be made carefully to ensure those correct inferences are drawn. We continue to think that weak theory is overrated [@ward:2016] and that machine learning and big data will allow us to learn new things. It will be interesting to see how the evidence for these claims is adjudicated with additional usage and careful evaluation. Social science has leaned on theory for (at least) seventy years, but it is time to find useful ways to incorporate recently developed methods in the arena of machine learning and modern statistics.

### 2020-11-10 paper/paper.tex
drop small edits into .Rmd

Although their analysis in the end does not support this point, we do not disagree that theory can important in predictive modeling. Of the various ways one could contrast ad hoc machine learning with ``theoretical'' modeling, B\&S focus essentially on theory as a mechanism for variable selection. In this sense, theory is in fact already used, informally, in nominally ad hoc predictive modeling related to conflict forecasting. The problem is that the extent of extant theories for a phenomenon like civil war is so broad and non-specific in terms. They do not provide much guidance on an exact specification of what variables and which functional form is required. As a result it is both not very clear what the ``correct'' model should be, nor easy to demonstrate that any particular ``theoretical'' model's ability to predict generalizes beyond the data that was used to create it in the first place. Rather than trying to contrast theory and machine learning as competing mechanisms for creating predictive models, it would be more productive to focus on how we can use the flexibility of machine learning models and the information predictive modeling may give us about our ability to ``explain'' a phenomenon to strengthen and whittle down the broad space of extant theories.

@@

Instantiating the theoretical model is more about choosing the correct variables. This is the contrast between theoretical prediction models and atheoretical machine learning that B\&S focus on. But this dichotomy is misguided. The data that are used in a predictive model have to come from somewhere. Even if the inclusion of some factors is not explicitly justified, it is hardly the case that one assembles data without the expectation that it will be useful. It is not a coincidence that nobody uses 1,180 variable strawman models whose variable vectors predominantly consist of 0's, even if such a model coincidentally ends up performing well. Rather, we posit that most of the cases B\&S might characterize as atheoretical enter the ``theoretical'', with some awareness of existing arguments and work that is used to select data sources for inclusion in a dataset.\footnote{While we do not undertake this, there are a variety of procedures that one might use for variable choices, ranging from staring at lists of variables to machine learning approaches.}

@@

Conversely, not all factors that are ``theoretically'' important or suggested by the literature are also important for prediction, and some factors that turn out to be good predictors do not originate from prior theory. For example, both Ward, Greenhill, and Bakke (2010) and Hill and Jones (2014) examine a variety of variables that are considered ``theoretically'' important for explaining civil war onset and state repression, respectively, and find that there is substantial variation in how useful they are for for actually predicting those outcomes. And, it is entirely possible that a focus on building accurate predictive models could uncover factors overlooked in the literature, e.g.~the strong association between infant mortality and a broad variety of conflict outcomes like political instability (Goldstone et al. 2010), irregular leadership changes (Beger et al. 2017), and coups (Beger and Ward 2020).

@@

Blair \& Sambanis have focused attention on comparing forecasting models and forecasts in the civil war domain. As social science becomes more adept at predictive analysis, this will doubtless be of increasing importance. However, these comparisons must be made carefully to ensure those correct inferences are drawn. We continue to think that weak theory is overrated (Ward 2016), and that machine learning and big data will allow us to learn new things. It will be interesting to see how the evidence for these claims is adjudicated with additional usage and careful evaluation. Social science has leaned on theory for (at least) seventy years, but it is time to find useful ways to incorporate recently developed methods in the arena of machine learning and modern statistics.

>>>>

Although their analysis in the end does not support this point, we do not disagree that theory can be important in predictive modeling. Of the various ways one could contrast ad hoc machine learning with ``theoretical'' modeling, B\&S focus essentially on theory as a mechanism for variable selection. In this sense, theory is in fact already used, informally, in nominally ad hoc predictive modeling related to conflict forecasting. The problem is that the extent of extant theories for a phenomenon like civil war is so broad and non-specific in terms. They do not provide much guidance on an exact specification of what variables and which functional form is required. As a result it is both not very clear what the ``correct'' model should be, nor easy to demonstrate that any particular ``theoretical'' model's ability to predict generalizes beyond the data that was used to create it in the first place. Rather than trying to contrast theory and machine learning as competing mechanisms for creating predictive models, it would be more productive to focus on how we can use the flexibility of machine learning models and the information predictive modeling may give us about our ability to ``explain'' a phenomenon to strengthen and whittle down the broad space of extant theories.

@@

Instantiating the theoretical model is more about choosing the correct variables. This is the contrast between theoretical prediction models and atheoretical machine learning that B\&S focus on. But this dichotomy is misguided. The data that are used in a predictive model have to come from somewhere. Even if the inclusion of some factors is not explicitly justified, it is hardly the case that one assembles data without the expectation that it will be useful. It is not a coincidence that nobody uses 1,180 variable strawman models whose variable vectors predominantly consist of 0's, even if such a model coincidentally ends up performing well. Rather, we posit that most of the cases B\&S might characterize as atheoretical enter the ``theoretical'' with some awareness of existing arguments and work that is used to select data sources for inclusion in a dataset.\footnote{While we do not undertake this, there are a variety of procedures that one might use for variable choices, ranging from staring at lists of variables to machine learning approaches.}

@@

Conversely, not all factors that are ``theoretically'' important or suggested by the literature are also important for prediction, and some factors that turn out to be good predictors do not originate from prior theory. For example, both Ward, Greenhill, and Bakke (2010) and Hill and Jones (2014) examine a variety of variables that are considered ``theoretically'' important for explaining civil war onset and state repression, respectively, and find that there is substantial variation in how useful they are for for actually predicting those outcomes. It is entirely possible that a focus on building accurate predictive models could uncover factors overlooked in the literature, e.g.~the strong association between infant mortality and a broad variety of conflict outcomes like political instability (Goldstone et al. 2010), irregular leadership changes (Beger et al. 2017), and coups (Beger and Ward 2020).

@@

Blair \& Sambanis have focused attention on comparing forecasting models and forecasts in the civil war domain. As social science becomes more adept at predictive analysis, this will doubtless be of increasing importance. However, these comparisons must be made carefully to ensure those correct inferences are drawn. We continue to think that weak theory is overrated (Ward 2016) and that machine learning and big data will allow us to learn new things. It will be interesting to see how the evidence for these claims is adjudicated with additional usage and careful evaluation. Social science has leaned on theory for (at least) seventy years, but it is time to find useful ways to incorporate recently developed methods in the arena of machine learning and modern statistics.
